---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# pbsEDM

[![Travis build status](https://travis-ci.org/luke-a-rogers/pbsEDM.svg?branch=master)](https://travis-ci.org/luke-a-rogers/pbsEDM)
[![Project Status: WIP – Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)
[![Coverage status](https://codecov.io/gh/luke-a-rogers/pbsEDM/branch/master/graph/badge.svg)](https://codecov.io/github/luke-a-rogers/pbsEDM?branch=master)

An R package to implement the delay embedding methods of Empirical Dynamic Modeling

## Installation

``` {r installation, eval = FALSE}
devtools::install_github("luke-a-rogers/pbsEDM")
```

## Functionality

TBA

## Examples

TBA

## Workflow During Development 
Both pushing to same repo, so always check on GitHub first, then (if necessary)
```
git fetch
git rebase
```

If get conflicts then see some notes at https://github.com/pacific-hake/hake-assessment (for which we are also pushing to a single repository), scroll down to the GitHub workflow section.
 
