---
title: "Analyse a Simple Time Series"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{simple_time_series}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6
)
```

```{r setup}
library(pbsEDM)
```

The tibble `Nx_lags_orig` is saved in the package and contains a simple time
series $N(t)$ as `Nt`, the lagged value $N(t-1)$ as `Ntmin1`, and first difference
$X(t) = N(t+1) - N(t)$ as `Xt` (and `Xtmin1` and `Xtmin2` the lagged values),
with EDM results for embedding dimension 2, using `rEDM` and Andy's original
function `EDM_pred_E_2()`. See `?Nx_lags_orig` for full details.

Using that as an example here, see `inclusion_issue.Rmd` vignette for comparison
of Andy's original results with results from `pbsEDM` and `rEDM`.

## Visualise data - can prob skip once new function done

Plot of data (before making movie):
```{r}
plotPanelMovie.df2(Nx_lags_orig,
                   only.final.plot=TRUE,
                   open.pdf=FALSE)
```

Movie -- see `inclusion_issue.Rmd` vignette.

The final panel shows the predicted vs observed from rEDM and from Andy's manual
calculations (not relevant here).

## Analyse with pbsEDM code

```{r pbsEDMcalc}
calc <- pbsEDM(Nx_lags_orig,
               lags = list(Xt = c(0:1))) # A list of results

# pred <- calc$xt_forecast

plot_observed(calc, late.num = 5)
```

```{r calc2}
# Use original Nt only, do lag of 1:
calc2 <- pbsEDM(Nx_lags_orig,
                lags = list(Nt = c(0:1)),
                first_difference = TRUE)

plot_observed(calc2, late.num = 5)

testthat::expect_equal(calc$xt_forecast, calc2$xt_forecast)
```

Loop over values of lag (then make into wrapper function):
```{r loop}
E_res = list()
par(mfrow = c(3,2))
for(E in 1:5){
  E_res[[E]] <- pbsEDM(Nx_lags_orig,
                       lags = list(Nt = 0:E),
                       first_difference = TRUE)
  plot_observed(E_res[[E]], late.num = 5)
}

```
