---
title: "Demonstration of Aspects 3 and 4"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{aspect_3_4}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "",
  fig.width = 5,
  fig.height = 5
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup}
library(pbsEDM)
```

## Introduction

In the manuscript, Aspect 3 is that predicted values of lagged variables can
give unrealistic negative predictions of the original unlagged populations.

Again we use the simulated population time series from the manuscript, with
population size denoted $N_t$ and first-differenced values denoted $Y_t$ at $t = 1, 2, 3, ..., 99, 100$. We
extract them from the saved tibble `NY_lags_example_3`:

```{r, t_and_Y}
input <-  dplyr::select(NY_lags_example_3, t, N_t, Y_t) %>%
  dplyr::rename(Time = t)
input
```

## Calculations using pbsEDM (as done in Aspect 2 vignette)

See the [`analyse_simple_time_series`](analyse_simple_time_series.html) vignette
for general details on using the code. Here do the simplex calculation for $E = 2$:
```{r pbsEDM}
pbsEDM_res_full <- pbsEDM::pbsEDM(input,
                                  lags = list(Y_t = c(0:1)))   # Gives full calculations

pbsEDM_pred_Y <- pbsEDM_res_full$X_forecast[-length(pbsEDM_res_full$X_forecast)]
                                           # Just the predictions, excluding
                                           #  t=101, so for t = 1:100
```

## Explaining the issue -- Aspect 3

Use the $Y_t$ calculations and then convert to $N_t$. Equation (1) in the
manuscript defines $Y_t$ as:

$Y_t = N_{t+1} - N_t$.

So having estimated $\hat{Y}_t$ we can rearrange that to estimate $\hat{N}_{t+1}$ as:

$\hat{N}_{t+1} = N_t + \hat{Y}_t$

or

$\hat{N}_t = N_{t-1} + \hat{Y}_{t-1}$,

to give a vector of estimated $\hat{N}_t$ values:
```{r Npred}
pbsEDM_pred_N = c(NA,
                  input$N_t + pbsEDM_pred_Y)
pbsEDM_pred_N
```

The first three `NA` values demonstrate that we cannot predict $\hat{N}_1$ (because we do not know $N_0$), or
$\hat{N}_2$ and $\hat{N}_3$ (because we cannot predict $\hat{Y}_1$ or
$\hat{Y}_2$).

Of note are the several negative predictions of $\hat{N}_t$, which
are unrealistic because our data represent populations -- this is Aspect 3. The indices of these
give the predicted time values (i.e. $t^* + 1$):
```{r indices}
which(pbsEDM_pred_N < 0)
```

Of particular interest is the negative prediction for $\hat{N}_{101}$ of
`r  pbsEDM_pred_N[101]`, since the
aim of the analysis is generally to forecast next year's population from all
currently available data.

## Simple solution to Aspect 3

The simplest solution to Aspect 3 is to just replace the negative
predictions $\hat{N}_t$ with the minimum observed value of $N_t$.
```{r replace}
N_t_min <- min(input$N_t)

pbsEDM_pred_N_replace <- pbsEDM_pred_N
pbsEDM_pred_N_replace[pbsEDM_pred_N_replace < 0] = N_t_min
```

Graphically this is shown by plotting all the original $(N_t, \hat{N}_t)$ points
as open black circles, then the set with the negative $\hat{N}_t$ values
replaced as red filled circles (so that 11 points can be seen to change; the 1:1
line is added in black):

```{r replacefig, echo=FALSE}
plot(c(input$N_t, NA),
     pbsEDM_pred_N,
     xlab = "Observed N_t",
     ylab = "Predicted N_t",
     main = "Red: change -ve predictions to min +ve observation")

points(c(input$N_t, NA),
       pbsEDM_pred_N_replace,
       pch = 20,
       col = "red")
abline(a = 0, b = 1)
# abline(h = mean_pred_N)
# abline(h = mean_pred_N_replace,
#        col = "red")
# abline(v = mean(input$N_t),
#        col = "grey")

# lm_fit  <- lm(pbsEDM_pred_N ~ c(input$N_t, NA))
# abline(lm_fit)
# lm_fit_replace  <- lm(pbsEDM_pred_N_replace ~ c(input$N_t, NA))
# abline(lm_fit_replace, col = "red")
```

## Aspect 4 - good fit for the first-differenced values ($Y_t$) does not necessarily translate to a good fit to the original population sizes ($N_t$)

The above simple fix solves Aspect 3 by removing unrealistic negative
predictions. However, the above figure raises a fundamental issue, in that the
predicted values $\hat{N}_t$ do not seem to match the original data $N_t$ too
well.

The correlation coefficient of $\hat{N}_t$ and $N_t$ before correcting the
negative values is:
```{r rhoN}
pbsEDM_rho_N <- cor(c(input$N_t, NA),
                    pbsEDM_pred_N,
                    use = "pairwise.complete.obs")
pbsEDM_rho_N
```
and after correcting the negative values it is
```{r replacerho}
pbsEDM_rho_N_replace <- cor(c(input$N_t, NA),
                            pbsEDM_pred_N_replace,
                            use = "pairwise.complete.obs")
pbsEDM_rho_N_replace    # cor(input$N_t[4:100], pbsEDM_pred_N_replace[4:100]) gives same
```

The latter $\rho$ with the negatives corrected is (unintuitively) very slightly lower
than the original value. But the more important point is that a $\rho$ of 0.28
is much lower than that based on $Y_t$ and $\hat{Y}_t$, which is
```{r rhoY}
pbsEDM_rho_Y <- pbsEDM_res_full$results$X_rho
pbsEDM_rho_Y
```

This seems to have not been documented before, and so we define it as Aspect 4.

Plotting  $\hat{Y}_t$ against $Y_t$ (these are the $E=2$ points in the
manuscript's Figure~1(f):

```{r Ycorr}
plot(input$Y_t,
     pbsEDM_pred_Y,
     xlab = "Observed Y",
     ylab = "Predicted Y")
abline(a=0, b=1)
```

shows a good correspondence, as given by the correlation coefficient of
`r pbsEDM_rho_Y`.

Thus, a good correspondence based on the first-differenced $Y_t$ values does not
necessarily lead to good correspondence in the $N_t$ values, which is what we
are interested in for practical situations. Hence our recommendation that
practitioners also check the fits of the predctions based on
non-first-differenced values ($N_t$) and not just first-differenced values
($Y_t$).

## Repeating analyses with $E=3$

The above was all for the simplest embedding dimension of $E=2$. We repeat the
calculations here for $E=3$, which is the optimal embedding dimension (based on
the maximum $\rho$ value), and so should be used for forecasting.
The code is available in the vignette code but not all
of it is echoed here. The first-differencing calculations remain the same, the higher embedding
dimension just allows more lags (and more neighbours) to be used in the making predictions.

```{r pbsEDM3}
pbsEDM_res_full_E_3 <- pbsEDM::pbsEDM(input,
                                      lags = list(Y_t = c(0:2)))   # Gives full calculations

pbsEDM_pred_Y_E_3 <- pbsEDM_res_full_E_3$X_forecast[-length(pbsEDM_res_full$X_forecast)]
                                           # Just the predictions, excluding
                                           #  t=101, so for t = 1:100
```

```{r Npred3}
pbsEDM_pred_N_E_3 = c(NA,
                  input$N_t + pbsEDM_pred_Y_E_3)
pbsEDM_pred_N_E_3
```

Interestingly, there are fewer negative predictions of $\hat{N}_t$ for $E=3$
than the 11 we had for $E=2$. The indices of these
give the predicted time values (i.e. $t^* + 1$):
```{r indices3}
which(pbsEDM_pred_N_E_3 < 0)
```

Though there is still a negative prediction for $\hat{N}_{101}$ of
`r  pbsEDM_pred_N_E_3[101]`.

Now solve Aspect 3 by replacing  the negative
predictions $\hat{N}_t$ with the minimum observed value of $N_t$.
```{r replace3}
N_t_min <- min(input$N_t)

pbsEDM_pred_N_E_3_replace <- pbsEDM_pred_N_E_3
pbsEDM_pred_N_E_3_replace[pbsEDM_pred_N_E_3_replace < 0] = N_t_min
```

And repeat the above figure:

```{r replacefig3, echo=FALSE}
plot(c(input$N_t, NA),
     pbsEDM_pred_N_E_3,
     xlab = "Observed N_t",
     ylab = "Predicted N_t",
     main = "Red: change -ve predictions to min +ve observation")

points(c(input$N_t, NA),
       pbsEDM_pred_N_E_3_replace,
       pch = 20,
       col = "red")
abline(a = 0, b = 1)
```

The above figure does look better (close to the 1:1 line) than for $E=2$.
The correlation coefficient of $\hat{N}_t$ and $N_t$ before correcting the
negative values is:
```{r rhoN3}
pbsEDM_rho_N_E_3 <- cor(c(input$N_t, NA),
                        pbsEDM_pred_N_E_3,
                        use = "pairwise.complete.obs")
pbsEDM_rho_N_E_3
```
and after correcting the negative values it is
```{r replacerho3}
pbsEDM_rho_N_E_3_replace <- cor(c(input$N_t, NA),
                                pbsEDM_pred_N_E_3_replace,
                                use = "pairwise.complete.obs")
pbsEDM_rho_N_E_3_replace
```

which again is slightly lower than the original value. But the $\rho$ has
improved somewhat from that for $E=2$, but ist still not as high as based on $Y_t$ and $\hat{Y}_t$, which is
```{r rhoY3}
pbsEDM_rho_Y_E_3 <- pbsEDM_res_full_E_3$results$X_rho
pbsEDM_rho_Y_E_3
```

So again, Aspect 4 occurs.

As above, plotting  $\hat{Y}_t$ against $Y_t$ (these are the $E=3$ points in the
manuscript's Figure~1(f):

```{r Ycorr3}
plot(input$Y_t,
     pbsEDM_pred_Y_E_3,
     xlab = "Observed Y",
     ylab = "Predicted Y")
abline(a=0, b=1)
```

shows a good correspondence, as given by the correlation coefficient of
`r pbsEDM_rho_Y_E_3`.

## Test that rEDM and pbsEDM results have not changed - DO SOMETHING SIMILAR

TODO maybe.

<!--
TODO

Here we run some tests to check that if any of the above results from `pbsEDM`
or `rEDM` have changed from when this vignette was written (some results are
saved in `NY_lags_example_3`). The tests will give errors if they fail (and any
such failures should be investigated and this vignette updated).

Check the default `rEDM` predictions have not changed.
{r testrEDM}
testthat::expect_equal(NY_lags_example_3$rEDM.pred,
                       rEDM_res$Predictions)       # If no error then they match


Also check that the results with different values of `exclusionsRadius` have not changed:
{r testExcl}
testthat::expect_equal(check_rEDM_excl_1, rEDM_pred_excl_1)
testthat::expect_equal(check_rEDM_excl_2, rEDM_pred_excl_2)
testthat::expect_equal(check_rEDM_excl_3, rEDM_pred_excl_3)


The test in the Independent Code section has already verified that the `pbsEDM`
results have not changed (since they match the `my.pred` results from the original
independent code).
-->
