---
title: "Inclusion Issue Version 2 (new rEDM to compare)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{inclusion_issue_2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 5,
  fig.height = 5
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup}
library(pbsEDM)
```

## Introduction

This vignette shows the code used to reproduce the issue we found with
calculations in the `rEDM` package, namely to do with the nearest neighbours
that are used in the projection calculations. We verified this using manual code written by Andrew Edwards and
then using functionalised code (in this `pbsEDM` package) written independently by
Luke Rogers. <!-- See our manuscript for a full explanation. -->

The tibble `NY_lags_example` is saved in the package and contains a simple
simulated population time
series $N_t$ as `Nt`, with $t = 1, 2, 3, ..., 99, 100$. It also contains the
lagged values $N_{t-1}$ as `Ntmin1`, and first differences $Y_t = N_{t+1} - N_t$
as `Y_t`, with lagged values $Y_{t-1}$ as `Y_tmin1` and $Y_{t-2}$ as `Y_tmin2`. Also
included are EDM results for embedding dimension 2, using `rEDM` (`rEDM.pred`
for the predicted value and `rEDM.var` for its variance) and Andy's original
function `EDM_pred_E_2()` (saved as `my.pred` and `my.var`). See `?NY_lags_example` for full details.

<!-- Change what's below based on
edm-work/code/simulated/sockeye-simulated/sockeye-sim-edm.Snw - has the movie
edm-work/code/simulated/egDeyle/egDeyle.rnw (that I used for different tstar,
for EAFMWG Jan 2020 I used egDeyleT95tstar15.rnw - can generalise once get
going here.
edm-work/code/simulated/egDeyle/tstarLoop/tstarLoop2019.rnw
and
edm-work/code/simulated/egDeyle/tstarLoop/mwe-2019.r
-->

Here are example results (for helping explain some of the later text):
```{r fullresults}
as.data.frame(head(NY_lags_example))
as.data.frame(tail(NY_lags_example))
```

Note that (TODO: check this):
```{r}
NY_lags_example[100, c("rEDM.pred", "my.pred")]
```
such that `rEDM` does not predict $Y_{100} = Y_{101} - Y_{100}$, but that is the
value we want so that we can predict $N_{101}$.
We should be able to predict it because we have all the relevent
information.

## Visualise data and some results

<!-- Plot of data (before making movie):
Commenting out since not really needed (and have better function now
demonstrated in the vignette analyse_simple_time_series, though the code here
compares results in the final panel.
```{r, eval=FALSE}
plotPanelMovie.df2(only.final.plot=TRUE,
                   open.pdf=FALSE)

# ```{r, animation.hook = 'gifski', interval = 1.5}
# Cannot change fig.width and fig.height hooks unless re-order par() in function
# fig.width is 0.67 * fig.height
#```{r, eval=FALSE}
 for(i in 1:nrow(NY_lags_example))
  {
    plotPanelMovie.df2(end = i,
                       max_time = max(NY_lags_example$t),
                       only.final.plot=TRUE,
                       open.pdf=FALSE)
  }
# ```
```
-->

Plot Andy's manual predictions against rEDM predictions, with the differing ones
in red:
```{r}
eps = 0.00001      # How different the predictions can be
plot(NY_lags_example$rEDM.pred,
     NY_lags_example$my.pred,
     xlab = "rEDM predictions",
     ylab = "Andy predictions")
abline(a=0,
       b=1,
       col="grey")
# Colour in red the ones more than eps away
different <- dplyr::filter(NY_lags_example,
                           abs(pred.diff) > eps)
points(different$rEDM.pred,
       different$my.pred,
       col = "red",
       pch = 20)

```

This shows that three points disagree between Andy's calculations and
rEDM's by more than $`r eps`$. (Originally there were a few more for other
reasons, but these have been corrected in rEDM). The three points are
```{r}
as.data.frame(different)
```

## Use new pbsEDM code

Now use the new `pbsEDM` code that Luke wrote independently; see the
[`analyse_simple_time_series`](analyse_simple_time_series.html) vignette for details on
using it. Then compare `pbsEDM` results with those from `rEDM` and from Andy's manual code:
```{r pbsEDMcalc}
pbs_calc <- pbsEDM(NY_lags_example,
                   lags = list(Y_t = c(0:1))) # A tibble (contains lists)

testthat::expect_equal(NY_lags_example$Y_t,
                       pbs_calc$X_observed[-length(pbs_calc$X_observed)]) # check the indexing is the same, X_observed now has extra value though
                       # TODO: check why that's needed, maybe not now?

pbs_pred <- pbs_calc$X_forecast[-length(pbs_calc$X_forecast)]

plot(NY_lags_example$rEDM.pred,
     pbs_pred,
     xlab = "rEDM predictions",
     ylab = "pbsEDM predictions")
abline(a=0,
       b=1,
       col="grey")
```

Looks similar to above figure (have not coloured in red points). Then compare
`pbsEDM` results with Andy's:
```{r}
plot(NY_lags_example$my.pred,
     pbs_pred,
     xlab = "Andy predictions",
     ylab = "pbsEDM predictions")
abline(a=0,
       b=1,
       col="grey")

# max(abs(NY_lags_example$my.pred - pbs_pred), na.rm = TRUE)
testthat::expect_equal(NY_lags_example$my.pred,
                       pbs_pred)
```

The above figure shows they look identical, as confirmed by the above line of code not
returning an error.

**Conclusion**: Andy's original manual calculations and Luke's new function in
`pbsEDM` give the same results for this simulated data set, which differs for
three time values from
results from `rEDM`. Andy's and Luke's code was written independently, although
Andy had explained the issue as to why results differed to those from `rEDM`.
This is a reassuring test of Luke's new code, which is more
general and functionalised than Andy's, and can do more than just $E=2$.

## Explaining the issue

Vector $\bf{x}_t$ is the vector of lagged values of the
scalars $Y_t$:
$$\bf{x}_t = [Y_t, ~Y_{t-1}].$$
For focal time $t^*$, we know $\bf{x}_{t^*}$ and are trying to predict
$\bf{x}_{t^*+1}$. The issue is that `rEDM` seems to use $\bf{x}_{t^*+2}$ as a
nearest neighbour, but this should not be permissible.

For example, for $t^*=94$, Andy's R code calculates nearest neighbours with indices $\psi$
(saved as `psivec94`) and corresponding weights `weights94`:
```{r psi}
psivec94 = c(6, 57, 88)
weights94 = c(0.3678794, 0.3205861, 0.2895013)
```
giving the estimate of $Y_{95}$, from [S1] of Deyle et al. (2013), as
```{r Yninetyfive}
Y95est = sum(weights94 * simple_ts[psivec94+1]) / sum(weights94)
Y95est
```

The `pbsEDM` code gives the same results:
```{r}
pbs_calc$neighbour_index[94, ]
pbs_calc$neighbour_weight[94, ]
pbs_calc$X_forecast[95]
```
The following will give errors if those results don't agree):
```{r test}
testthat::expect_equal(psivec94, pbs_calc$neighbour_index[94, ])
testthat::expect_equal(weights94, pbs_calc$neighbour_weight[94, ], tolerance = 0.000001)
testthat::expect_equal(Y95est, pbs_calc$X_forecast[95])
```
However, `rEDM` gives
```{r rEDMninetyfive}
Y95est_rEDM = rEDM_points[95, "pred"]
Y95est_rEDM
```

TODO: redoing that and check if it's changed.

## Manually reproduce `rEDM` result

We can reproduce the `rEDM` result by adapting Andy's manual code to allow
$\bf{x}_{96}$ to be a nearest neighbour of $\bf{x}_{94}$,  giving
```{r psi94allow}
psivec94_allow = c(96, 6, 57)   # so 6 and 57 are now 2nd and 3rd nearest neighbours
weights94_allow = c(3.678794e-01, 1.405278e-04, 4.146457e-05)
# Note that the first weight is the same as above (by definition it's always
#  exp(-1)), but the second and third are very small because x[96] is
#  very close to x[94].
Y95est_allow = sum(weights94_allow * simple_ts[psivec94_allow+1]) /
                   sum(weights94_allow)
Y95est_allow
```

However, the problem is that we should not be allowed to use $\bf{x}_{96}$.
By definition,
$$\bf{x}_{96} = [Y_{96}, Y_{95}],$$
which contains $Y_{95}$. But $Y_{95}$ is the first component of
$$\bf{x}_{95} = [Y_{95}, Y_{94}]$$
and is exactly what we are trying to predict.
$Y_{95}$ is presumably the 'one' that we are leaving out in 'leave-one-out', but
it occurs in $\bf{x}_{95}$ and $\bf{x}_{96}$, by definition.

So $\bf{x}_{96}$ should not be used to forecast $\bf{x}_{95}$, because it
explicitly contains one of the two components of $\bf{x}_{95}$.

The same issue occurs (with Andy's manual code) for $t^*=75$, for which Luke's code
gives the same answer as shown above (because all values give the same answer).

To summarise -- the issue is that rEDM does not fully take into account that
$\bf{x}_t$ is a vector of lagged values.
<!-- Not sure what I meant by this: It's not simply an implementation issue
(that `rEDM` is unaware that the $X_t$ are differences of the $N_t$). -->

## Another attempt with `rEDM`

Andy tried the following in `rEDM`, to not allow $Y_{96}$ to be a nearest neighbour
predicting $Y_{95}$ from $Y_{94}$ -- but think this isn't possible, since it still
kind of wants cross-validation. Had played with `lib` and `pred` but couldn't get answer
to agree with Andy's original one. Not running
this in vignette, keeping for reference if needed:
```{r rEDMagain, eval=FALSE}
simp_Efix2 = simplex(simple_ts,
                     E = 2,
                     stats_only = FALSE,
                     # lib = matrix(c(1, 94, 97, 99), nrow = 2, byrow = TRUE),
                     lib = c(1, 94),
                     pred = c(95, 97)
                     # pred = matrix(c(95, 95, 96, 96), nrow = 2, byrow = TRUE)
                     )
rEDM_points2 = simp_Efix2[,"model_output"][[1]]       # time, obs, pred and pred_var
# Make each row correspond to t:
# rEDM_points2 = rbind(c(1, simple_ts[1], NA, NA), rEDM_points2) - only if doing all times
head(rEDM_points2)
```
